Predict Disease Outcome by Community Structures

### 2016-09-07 code/HumanCdGF_entero_predict.Rmd
update predict files to work with latest meta file

meta_file   <- read.table(meta_file, sep = '\t', header = T)
shared_file <- read.table(shared_file, sep='\t',header = T, row.names = 'Group')
tax_file <- read.table(tax_file, sep='\t',header = T, row.names = 'OTU')
entero_list <- read.table(entero_list, sep = '\t', header = T)
entero_abund <- read.table(entero_abund, sep = '\t', header = T)

@@

# identify c difficile otus
cdiff_otu <- rownames(tax_file[grep('Clostridium_XI', tax_file$Taxonomy),])
# remove samples not used or with other cdiff strains
meta_file <- 
  meta_file %>% 
  filter(sample_type == 'stool', cdiff_strain == 431) %>% 
  select(Mouse_ID, cage_id, group, day, cdiff_cfu, Euthanized) %>% 
  mutate(log_cfu = log10(cdiff_cfu + 1)) %>% 
  filter(!is.na(log_cfu))

>>>>

meta_file   <- read.table(meta_file, sep = '\t', header = T)
shared_file <- read.table(shared_file, sep='\t',header = T, row.names = 'sample_id')
tax_file <- read.table(tax_file, sep='\t',header = T, row.names = 'OTU')
entero_list <- read.table(entero_list, sep = '\t', header = T)
entero_abund <- read.table(entero_abund, sep = '\t', header = T)

@@

# identify c difficile otus
cdiff_otu <- rownames(tax_file[grep('Clostridium_XI', tax_file$Taxonomy),])
# remove samples not used or with other cdiff strains
meta_file <- 
  meta_file %>% 
  filter(cdiff_strain == 431) %>% 
  select(mouse_id, cage_id, group, day, cdiff_cfu, Euthanized) %>% 
  filter(!is.na(log_cfu))

### 2016-09-07 code/HumanCdGF_microbiota_predict.Rmd
update predict files to work with latest meta file

setwd("~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code")

# read in files
meta_file   <- read.table('../data/process/human_CdGF_metadata.txt', sep = '\t',
                        header = T, row.names = 'group')
shared_file <- read.table('../data/process/human_CdGF.an.unique_list.0.03.subsample.shared',
                          sep='\t',header = T, row.names='Group')
tax_file <- read.table('../data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy',
                       sep='\t',header = T, row.names = 'OTU')
source('tax_level.R')

# add log transformed CFU
meta_file$log_cfu <- log10(meta_file$cdiff_cfu + 1)
# remove cecum samples
meta_file <- meta_file[meta_file$sample_type == 'stool', ]
# remove unused cage P2 and 
meta_file <- meta_file[! meta_file$cage_id %in% 'P2', ]
# select only mice with 431 cdiff
meta_file <- meta_file[meta_file$cdiff_strain == '431', ]

# subset shared
#Create df with relative abundances
stool_shared <- shared_file[rownames(meta_file[meta_file$day %in% 0, ]) ,]
# use %in% instead of == because to ignore NA
# rename day 0 community rows by mouse id - since meta rows have day in row names, use mouse id for both
rownames(stool_shared) <- meta_file$Mouse_ID[rownames(meta_file) %in% 
                                               rownames(stool_shared) ]
rel_abund_stool <- 100 * stool_shared / unique(apply(stool_shared, 1, sum))
#Create vector of OTUs with median abundances >1%
OTUs_stool_0_01 <- apply(rel_abund_stool, 2, max) > 1
#df of OTUs with abundances >1% - by cage and inoculum
rel_abund_stool <- rel_abund_stool[ ,OTUs_stool_0_01]
OTU_list <- colnames(rel_abund_stool)

@@

#log_cfu_obs <- meta_file[meta_file$day %in% c(0:10), c('Mouse_ID', 'day', 'log_cfu') ]
#log_cfu_obs$predict <- NA
#log_cfu_obs[log_cfu_obs$day==2, ] <- predict(RF, rel_abund_stool)

@@

for (i in 1:10){
  predict_all_df <- merge(meta_file[meta_file$day==i, c('Mouse_ID', 'day' , 'log_cfu') ],
                    rel_abund_stool, by='Mouse_ID', by.y='row.names', all.y = T)
  predict_df <- predict_all_df[,! colnames(predict_all_df) %in% 'day']
  predict_df <- predict_df[ ,!(colnames(predict_df) %in% "Mouse_ID")]
  RF <- randomForest(log_cfu ~ ., data = predict_df[!is.na(predict_df$log_cfu), ],
                   importance = TRUE, ntree = n_trees)
  rsq_df[i] <- RF$rsq[length(RF$rsq)]
  mse_df[i] <- RF$mse[length(RF$mse)]

@@

CFU_euth <- merge(meta_file[meta_file$day==meta_file$Euthanized, c('Mouse_ID', 'log_cfu') ],
                    rel_abund_stool, by='Mouse_ID', by.y='row.names', all.y = T)
CFU_euth <- CFU_euth[ ,!(colnames(CFU_euth) %in% "Mouse_ID")]

@@

without_early_euth <- meta_file[meta_file$Euthanized %in% 10 ,c('day', 'Mouse_ID', 'log_cfu')]
without_early_euth <- without_early_euth[without_early_euth$day %in% c(1:4), ]

without_by_day <- merge(without_early_euth, shared_X_cdiff_ra, by  = 'row.names')
without_day_0 <- merge(without_early_euth, rel_abund_stool, by='Mouse_ID', by.y='row.names' )

@@

for (i in 1:4){
  predict_all_df <- without_day_0[without_day_0$day == i ,]
  predict_df <- predict_all_df[!is.na(predict_all_df$log_cfu), 
                           !colnames(predict_all_df) %in% c('Mouse_ID', 'day')]
  RF <- randomForest(log_cfu ~ ., data = predict_df,
                   importance = TRUE, ntree = n_trees)
  rsq_persist[i] <- RF$rsq[length(RF$rsq)]
  mse_persist[i] <- RF$mse[length(RF$mse)]

@@

for (i in 1:4){
  predict_all_df <- without_by_day[without_by_day$day == i ,]
  predict_df <- predict_all_df[!is.na(predict_all_df$log_cfu), 
                           !colnames(predict_all_df) %in% c('Mouse_ID', 'Row.names', 'day')]
  RF <- randomForest(log_cfu ~ ., data = predict_df,
                   importance = TRUE, ntree = n_trees)
  rsq_persist_day[i] <- RF$rsq[length(RF$rsq)]
  mse_persist_day[i] <- RF$mse[length(RF$mse)]

@@

predict_cfu_persist_df <- without_by_day[!is.na(without_by_day$log_cfu),
                                           ! colnames(without_by_day) %in% c('Row.names','day','Mouse_ID')]
RF_cfu_persist_rf <- randomForest(log_cfu ~ ., data = predict_cfu_persist_df,
                   importance = TRUE, ntree = n_trees)
RF_cfu_persist_df <- data.frame(cbind(without_by_day[!is.na(without_by_day$log_cfu), 
                                                       c('day','log_cfu')], 
                                  predicted = RF_cfu_persist_rf$predicted))

@@

##Can we predict moribundity with the community structure at Day 0?
#### randomForest
```{r fig.width=10, fig.height=5}
meta_file$Euth_Early <- factor(ifelse(meta_file[ , 'Euthanized'] %in% 10, 1, 0))

Predict_early_euth_df <- merge(meta_file[meta_file$day == 0, c('Mouse_ID', 'Euth_Early')], 
                               rel_abund_stool, by.x = 'Mouse_ID', by.y='row.names')
Predict_early_euth_df_names <- Predict_early_euth_df$Mouse_ID
Predict_early_euth_df <- select(Predict_early_euth_df, -Mouse_ID)

>>>>

meta_file   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.cons.taxonomy'
tax_function <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R'

# read in files
meta_file   <- read.table(meta_file, sep = '\t', header = T, row.names = 'sample_id')
shared_file <- read.table(shared_file, sep = '\t', header = T, row.names = 'sample_id')
tax_file <- read.table(tax_file, sep = '\t', header = T, row.names = 'OTU')
source(tax_function)

# select only mice with 431 cdiff
meta_file <- meta_file[meta_file$cdiff_strain == '431', ]

# subset shared
#Create df with relative abundances
stool_shared <- shared_file[rownames(meta_file[meta_file$day %in% 0, ]) ,]
# use %in% instead of == because to ignore NA
# rename day 0 community rows by mouse id - since meta rows have day in row names, use mouse id for both
rownames(stool_shared) <- meta_file$mouse_id[rownames(meta_file) %in% 
                                               rownames(stool_shared) ]
rel_abund_stool <- 100 * stool_shared / unique(apply(stool_shared, 1, sum))
#Create vector of OTUs with median abundances >1%
OTUs_stool_0_01 <- apply(rel_abund_stool, 2, max) > 1
#df of OTUs with abundances >1% - by cage and inoculum
rel_abund_stool <- rel_abund_stool[ ,OTUs_stool_0_01]
OTU_list <- colnames(rel_abund_stool)

@@

#log_cfu_obs <- meta_file[meta_file$day %in% c(0:10), c('mouse_id', 'day', 'log_cfu') ]
#log_cfu_obs$predict <- NA
#log_cfu_obs[log_cfu_obs$day==2, ] <- predict(RF, rel_abund_stool)

@@

for (i in 1:10){
  predict_all_df <- merge(meta_file[meta_file$day==i, c('mouse_id', 'day' , 'log_cfu') ],
                    rel_abund_stool, by='mouse_id', by.y='row.names', all.y = T)
  predict_df <- predict_all_df[,! colnames(predict_all_df) %in% 'day']
  predict_df <- predict_df[ ,!(colnames(predict_df) %in% "mouse_id")]
  RF <- randomForest(log_cfu ~ ., data = predict_df[!is.na(predict_df$log_cfu), ],
                   importance = TRUE, ntree = n_trees)
  rsq_df[i] <- RF$rsq[length(RF$rsq)]
  mse_df[i] <- RF$mse[length(RF$mse)]

@@

CFU_euth <- merge(meta_file[meta_file$day==meta_file$last_day_by_mouse, c('mouse_id', 'log_cfu') ],
                    rel_abund_stool, by='mouse_id', by.y='row.names', all.y = T)
CFU_euth <- CFU_euth[ ,!(colnames(CFU_euth) %in% "mouse_id")]

@@

without_early_euth <- meta_file[meta_file$Early_Euth == FALSE ,c('day', 'mouse_id', 'log_cfu')]
without_early_euth <- without_early_euth[without_early_euth$day %in% c(1:4), ]

without_by_day <- merge(without_early_euth, shared_X_cdiff_ra, by  = 'row.names')
without_day_0 <- merge(without_early_euth, rel_abund_stool, by='mouse_id', by.y='row.names' )

@@

for (i in 1:4){
  predict_all_df <- without_day_0[without_day_0$day == i ,]
  predict_df <- predict_all_df[!is.na(predict_all_df$log_cfu), 
                           !colnames(predict_all_df) %in% c('mouse_id', 'day')]
  RF <- randomForest(log_cfu ~ ., data = predict_df,
                   importance = TRUE, ntree = n_trees)
  rsq_persist[i] <- RF$rsq[length(RF$rsq)]
  mse_persist[i] <- RF$mse[length(RF$mse)]

@@

for (i in 1:4){
  predict_all_df <- without_by_day[without_by_day$day == i ,]
  predict_df <- predict_all_df[!is.na(predict_all_df$log_cfu), 
                           !colnames(predict_all_df) %in% c('mouse_id', 'Row.names', 'day')]
  RF <- randomForest(log_cfu ~ ., data = predict_df,
                   importance = TRUE, ntree = n_trees)
  rsq_persist_day[i] <- RF$rsq[length(RF$rsq)]
  mse_persist_day[i] <- RF$mse[length(RF$mse)]

@@

predict_cfu_persist_df <- without_by_day[!is.na(without_by_day$log_cfu),
                                           ! colnames(without_by_day) %in% c('Row.names','day','mouse_id')]
RF_cfu_persist_rf <- randomForest(log_cfu ~ ., data = predict_cfu_persist_df,
                   importance = TRUE, ntree = n_trees)
RF_cfu_persist_df <- data.frame(cbind(without_by_day[!is.na(without_by_day$log_cfu), 
                                                       c('day','log_cfu')], 
                                  predicted = RF_cfu_persist_rf$predicted))

@@

##Can we predict moribundity with the community structure at Day 0?
#### randomForest
```{r fig.width=10, fig.height=5}
meta_file$Euth_Early <- factor(ifelse(meta_file[ , 'Early_Euth'] == FALSE, 1, 0))

Predict_early_euth_df <- merge(meta_file[meta_file$day == 0, c('mouse_id', 'Euth_Early')], 
                               rel_abund_stool, by.x = 'mouse_id', by.y='row.names')
Predict_early_euth_df_names <- Predict_early_euth_df$mouse_id
Predict_early_euth_df <- select(Predict_early_euth_df, -mouse_id)
